---
title: "Project Report 1: Exploring Gender Diveristy in Tech"
author: "Isobel Bodefeld"
date: "`r Sys.Date()`"
output: html_document
---

**Diversity in Tech**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
```


```{r}
# load in datasets

# survey results from a Hacker News survey of user salaries in the tech industry in 2016, providing detailed job titles, experience years, geographical locations, annual benefits, and signing and stock bonuses from employers
salaries_survey <- read_csv("salaries_clean.csv")

# workforce distributions by job category and race for 177 of the largest tech companies headquartered in Silicon Valley
raw_silicon <- read_csv("Reveal_EEO1_for_2016.csv")
```

**Introduction**
In this project, I aim to explore diversity in the technology industry. I was drawn to this topic because of my career interest in technology/data science and hope to learn more about the field while applying my data science skill set. I expect to see higher salaries and positions for males as compared to their female counterparts of other races, which tends to be the stereotype of the industry.

Research question: Does company gender proportion differ by annual base pay and/or by state? 

The first data set is comprised of survey results from a Hacker News survey of user salaries in the tech industry in 2016, providing detailed job titles, experience years, geographical locations, annual benefits, and signing and stock bonuses from employers. Each row contains a unique survey entry/ participant and there are both numeric and categorical variables.  

The second data set is of demographics of the employees from 25 companies in Silicon Valley with data from EEO-1 reports, including gender, race, and job category. 

Both data sets were found through Kaggle. I plan to join them using the company names found in each set, called employer_name and company.

sources: 

Brandon Telle. 2016. Retrieved October 1, 2023. https://www.kaggle.com/datasets/thedevastator/know-your-worth-tech-salaries-in-2016  

RACHAEL TATMAN. 2017. Retrieved October 1, 2023. https://www.kaggle.com/datasets/rtatman/silicon-valley-diversity-data/versions/1/data



Limitations to this analysis include: only male or female is listed as options for gender in EEO-1 reports; the Silicon Valley data is from 2016 while the US census data is from 2013; and I don't have data on who applied to these companies.


```{r}
# look at top 6 values for each dataset
head(salaries_survey)
head(raw_silicon)
```

**Wrangling/Tidying raw_silicon Before Joining/Merging**

```{r}
# convert the "count" column to numeric
raw_silicon$count <- as.numeric(raw_silicon$count)

# remove rows with missing values (N/A) in the "count" column
raw_silicon <- raw_silicon %>% na.omit()

# create a new data frame by grouping and summarizing the data
company_gender <- raw_silicon %>%
  group_by(company, gender) %>%
  summarize(count = sum(count)) %>%
  ungroup()

# extract male data and rename columns
silicon_demographics <- company_gender %>% filter(gender == "male") %>%
  select(-gender) %>%
  rename(male_count = count)

# extract female data and rename columns
femaleData <- company_gender %>% filter(gender == "female") %>%
  rename(female_count = count)

# calculate male proportion
silicon_demographics <- silicon_demographics %>%
  mutate(female_count = femaleData$female_count) %>%
  mutate(male_proportion = male_count / (male_count + female_count))

# calculate female proportion
silicon_demographics <- silicon_demographics %>%
  mutate(female_proportion = 1 - male_proportion)

# convert the "company" column to lowercase
silicon_demographics <- silicon_demographics %>%
  mutate(company = str_to_lower(company)) 

head(silicon_demographics)
```

The original data set raw_silicon that showed demographics of different companies had 3640 observations. Now that these functions were performed to show male and female proportions, the new data set silicon_demographics has 22 observations. This tidies the data and makes each row contain a unique company.

**Joining/Merging**

```{r}
# join datasets using employer_name and company as ID variables
combined_temp <- inner_join(salaries_survey, silicon_demographics, by = c("employer_name" = "company"))
head(combined_temp)

# count number of IDs the data sets have in common
num_combined_temp <- nrow(combined_temp)
num_combined_temp
```

Using inner_join, only the ID variables common to both data sets are kept. I use the variable "employer_name" from the survey data and the variable "company" from the company demographics data. They have overlapping companies but have different columns names so I combined the variable names in order to join effectively. There are 194 IDs that each data set have in common.

```{r}
# count observations in salaries_survey data
total_obs_salaries_survey <- nrow(salaries_survey)
total_obs_salaries_survey

# count observations in silicon_demographics data
total_obs_silicon_demographics <- nrow(silicon_demographics)
total_obs_silicon_demographics
```

Before joining, salaries_survey had 1655 observations and silicon_demographics had 22 observations. 

```{r}
# IDs in salaries_survey data but not silicon_demographics data
missing_IDs_salaries_survey <- salaries_survey %>%
  anti_join(silicon_demographics, by = c("employer_name" = "company"))
missing_IDs_salaries_survey

num_missing_IDs_salaries_survey <- nrow(missing_IDs_salaries_survey)
num_missing_IDs_salaries_survey

# IDs in silicon_demographics data but not salaries_survey data
missing_IDs_silicon_demographics <- silicon_demographics %>%
  anti_join(salaries_survey, by = c("company" = "employer_name"))
missing_IDs_silicon_demographics

num_missing_IDs_silicon_demographics <- nrow(missing_IDs_silicon_demographics)
num_missing_IDs_silicon_demographics
```

1461 IDs appeared in the salaries_survey data but not the silicon_demographics data. 4 IDs appeared in the silicon_demographics data but not the salaries_survey data. The observations that were dropped while combining are above.

There may be possible issues in not being able to explore the diversity of all companies listed from the survey, but the company demographics data limits exploration to the companies present within it.

Note: I exported the combined data set to Excel and manually filled out the location_state and location_country columns.

```{r}
# loading in updated combined data set with states and countries
combined <- read_csv("combined_updated.csv")
head(combined)
```

**Wrangling Merged Data**
```{r}
# calculate summary statistics for annual base pay
summary_annual_base_pay <- combined %>%
  summarise(
    Mean_Annual_Pay = mean(annual_base_pay, na.rm = TRUE),
    Median_Annual_Pay = median(annual_base_pay, na.rm = TRUE),
    Min_Annual_Pay = min(annual_base_pay, na.rm = TRUE),
    Max_Annual_Pay = max(annual_base_pay, na.rm = TRUE)
  )
summary_annual_base_pay

# calculate summary statistics for total experience years
summary_total_experience_years <- combined %>%
  summarise(
    Mean_total_experience_years = mean(total_experience_years, na.rm = TRUE),
    Median_total_experience_years = median(total_experience_years, na.rm = TRUE),
    Min_total_experience_years = min(total_experience_years, na.rm = TRUE),
    Max_total_experience_yearsy = max(total_experience_years, na.rm = TRUE)
  )
summary_total_experience_years

# calculate summary statistics for job title category
summary_job_title <- combined %>%
  group_by(job_title_category) %>%
  summarise(
    Count = n(),
    Unique_Count = n_distinct(employer_name),
    Median_Annual_Pay = median(annual_base_pay, na.rm = TRUE),
    Mean_Annual_Pay = mean(annual_base_pay, na.rm = TRUE)
  )
summary_job_title
```

After calculating summary statistics for annual base pay, I found the mean of $217106.40 to be significantly higher than the median of $131,000. This means there are more high annual pay outliers drawing the mean up (lower outliers have less of an effect). Because the minimum and maximum are drastically different from each other, I would opt to use the median when describing the center of the data. 

For the summary statistics of total experience years, I found the mean to be 6 years and the median to be 4 years. With a minimum and 0 years and maximum of 30 years, the statistics that describe the center suggest that the participants tend to be less far along in their career which just a few years of experience.

As for job title category, the majority of participants were in software with the count be 137 participants and the next highest being 21. The median annual pays are all between $110,000 and $170,000 whereas the mean annual pays had some higher values. In management, the highest mean was $1,190,604 which suggests high outlier(s) that draw this value up. For both median and mean, management was the job title category with the highest value whereas engineering had the lowest value.

**Visualizing**

```{r}
# visualization with 1 variable

# define the bin widths
pay_ranges <- c(0, 50000, 75000, 100000, 125000, 150000, 175000, 200000, Inf)

# add a factor variable with labels for the bins using mutate
combined_data_1 <- combined %>% 
  mutate(annual_base_pay_ranges = cut(annual_base_pay, breaks = pay_ranges, labels = c("0-50k", "50-75k", "75-100k", "100-125k", "125-150k", "150-175k", "175-200k", "200K+")))

# create a histogram
combined_data_1 %>% 
  ggplot(aes(x = annual_base_pay_ranges)) +
  geom_bar(fill = "#CC5500") +
  labs(x = "Annual Base Pay", y = "Count") +
  ggtitle("Distribution of Annual Base Pay") + 
  theme_classic()
```

This barplot shows the distribution of the amount of participants that fall into each of the annual base pay ranges. It is easy to see that the $100-125k, $125-150k and $150-175k annual base pay ranges are the top 3, respectively, highest ranges for number of participants that earn that much. The $50-75k range has the lowest amount of participants. I should point out that some observations contain interns rather than full-time employees, so there are more earning less that $50k.

```{r}
# visualization with 2 variables

# define custom pay ranges
pay_ranges <- c(0, 50000, 75000, 100000, 125000, 150000, 175000, 200000, Inf)

# Create a new variable "pay_range" based on the custom ranges
combined_data_2 <-combined %>%
  mutate(pay_range = cut(annual_base_pay, breaks = pay_ranges, labels = c("0-50k", "50-75k", "75-100k", "100-125k", "125-150k", "150-175k", "175-200k", "200K+"), right = FALSE))

# grouped bar plot with companies as groups and states as bars within each group
ggplot(combined_data_2, aes(x = pay_range, y = male_proportion, fill = pay_range)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Gender Proportion by Annual Base Pay Range", x = "Annual Base Pay Range", y = "Male Proportion") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme_classic()
```

In this visualization, I add male proportion as another variables so I can explore how company gender proportion varies by annual base pay. Through this barplot we can see that the $100-125k and $125-150k ranges have the highest proportion of males at their companies. However, there are also more observations for these ranges so it is hard to make definitive conclusions.

```{r}
# visualization with 3 variables

# define custom pay ranges
pay_ranges <- c(0, 50000, 75000, 100000, 125000, 150000, 200000, Inf)

# create new variable "pay_range" based on the custom ranges
combined_data_3 <- combined %>%
  filter(!is.na(location_state)) %>%
  mutate(pay_range = cut(annual_base_pay, breaks = pay_ranges, labels = c("0-50k", "50-75k", "75-100k", "100-125k", "125-150k", "150-200k", "200K+"), right = FALSE))

# grouped bar plot by pay range and state
ggplot(combined_data_3, aes(x = pay_range, y = male_proportion, fill = location_state)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Gender Proportion by Annual Base Pay Range and State",
       x = "Annual Base Pay Range", y = "Male Proportion") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme_classic()
```

By adding state, I aim to explore how gender proportion varies both by annual base pay and by state. For many of the pay ranges, California had the highest proportion of males. In addition, there is no clear relationship between male proportion and annual base pay nor state. It is notable to point out that a majority of the observations were of participants who work in California.  
